### 2016-05-29 text/Chapter07.tex
Addressing feedback, fix demands (Chapter 6)

\begin{example}[Flat dataflow]
In a context-aware language for dataflow (and in language with liveness tracking), the inherent
ambiguity of the (\emph{abs}) rule is resolved by placing the context requirements of the body on
both the declaration site and the call site. In Section~\ref{sec:flat-unique}, this was defined by
replacing the standard coeffect (\emph{abs}) rule with a rule (\emph{idabs}) that uses an
annotation $\cclrd{r}$ for the body of the function, declaration site coeffect and call site coeffect.

@@

\vspace{-0.5em}
\paragraph{Flat and structural dataflow.}
In flat dataflow, the context requirements of the body is required from both the declaration site
and from the call site. In structural dataflow, the context requirements are tracked separately
for each variable, which provides a more precise type. Consider the following two examples (the
\kvd{let} keyword is used to define a curried function of two arguments):
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{oldy}~\ident{x}~\ident{y}~=~\ident{x}~+~\kvd{prev}~\ident{y}~\kvd{in}\\[-0.25em]
\ident{oldy}
\end{array}
\end{equation*}
%
When type checking the expression using the flat system, the type of \ident{oldy} is inferred as {\tt num -\{1\}-> num -\{1\}-> num},
but when using the structural system, the type becomes {\tt num -\{0\}-> num -\{1\}-> num}.

>>>>

\begin{example}[Flat dataflow]
In a context-aware language for dataflow (and in language with liveness tracking), the inherent
ambiguity of the (\emph{abs}) rule is resolved by placing the context demands of the body on
both the declaration site and the call site. In Section~\ref{sec:flat-unique}, this was defined by
replacing the standard coeffect (\emph{abs}) rule with a rule (\emph{idabs}) that uses an
annotation $\cclrd{r}$ for the body of the function, declaration site coeffect and call site coeffect.

@@

\vspace{-0.5em}
\paragraph{Flat and structural dataflow.}
In flat dataflow, the context demands of the body is required from both the declaration site
and from the call site. In structural dataflow, the context demands are tracked separately
for each variable, which provides a more precise type. Consider the following two examples (the
\kvd{let} keyword is used to define a curried function of two arguments):
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{oldy}~\ident{x}~\ident{y}~=~\ident{x}~+~\kvd{prev}~\ident{y}~\kvd{in}\\[-0.25em]
\ident{oldy}
\end{array}
\end{equation*}
%
When type checking the expression using the flat system, the type of \ident{oldy} is inferred as {\tt num -\{1\}-> num -\{1\}-> num},
but when using the structural system, the type becomes {\tt num -\{0\}-> num -\{1\}-> num}.

### 2016-05-29 text/Chapter08.tex
Addressing feedback, fix demands (Chapter 6)

\item In the flat coeffect calculus, we distinguish between \emph{splitting} context requirements
  and \emph{merging} of context requirements ($\cpar$ and $\czip$, respectively). In the structural
  coeffect calculus, the operations (which model appending and splitting vectors) are invertible
  and so the structural coeffect algebra uses the same tensor product $\atimes$. However, the
  $\cpar$ operation is still needed in the contraction rule.
\end{itemize}

\noindent
In the unified calculus presented in this section, we address the three differences as follows.
We include explicit rules for context manipulation in the calculus; in systems that arise from flat
calculi, the structural rules do not change coeffects and can thus be applied freely. We generalize
the structure of coeffect annotations using the notion of a ``container'' which can be specialized
to obtain a single annotation or a vector of annotations. This could potentially be used to capture
other structures such as trees in bunched typing \cite{substruct-bunched}. Finally, we distinguish
between splitting and merging of context requirements (using the notation $\apar$ and $\azip$,
respectively). For structural coeffect calculi, the two operators coincide, but for flat
coeffect calculi, they differ and provide the needed flexibility.

@@

The $\cpar$ operation models combining of context requirements arising from multiple parts of a
program. The meaning depends on the coeffect container. The operation can either combine requirements
of individual variables (structural coeffects) or requirements attached to the whole context of
multiple sub-exp\-ressions (flat coeffects).

@@

The difference between $\azip$ and $\apar$ is clarified by the semantics (Sec\-tion~\ref{sec:further-semantics}),
where $\aclrd{\textbf{r}}\,\azip\,\aclrd{\textbf{s}}$ is an annotation of the \emph{codomain} of
a morphism that merges the capabilities provided by two contexts (in the syntactic
reading, splits the context requirements), while $\aclrd{\textbf{r}}\,\apar\,\aclrd{\textbf{s}}$
is an annotation of the \emph{domain} of a morphism that splits the capabilities of a single
context into two parts (in the syntactic reading, merges their context requirements).
Syntactically, this means that we always use $\azip$ in the rule \emph{assumptions} and
$\apar$ in \emph{conclusions}.

@@

In the (\emph{app}), (\emph{abs}) and (\emph{let}) rules, the only change from the structural
system is that the vector concatenation $\atimes$ is now replaced with context splitting/merging
of the unified coeffect algebra. As already mentioned, we use splitting of context requirements $\azip$
in rule \emph{assumptions} and merging of context requirements $\apar$ in rule \emph{conclusions}.

@@

\noindent
Intuitively, the requirement $(\cclrd{r}\,\czip\,\cclrd{s}) \;\cleq\; (\cclrd{r}\,\cpar\,\cclrd{s})$,
which could be also written as $(\aclrd{r}\,\azip\,\aclrd{s}) \;\aclrd{\leq}\; (\aclrd{r}\,\apar\,\aclrd{s})$,
denotes that splitting context requirements and then re-combining them preserves all the requirements
from the assumptions. The system may be imprecise and conclusions can overapproximate assumptions, but
it cannot lose requirements. This is fundamental for showing that exchange and contraction are admissible
in the unified system.

@@

\begin{itemize}
\item \textsc{Coeffect interpretation.} The \ident{counit} operation extracts a value
  and does not require any additional context; the \ident{cobind} operation requires
  context $\cclrd{r}\,\cseq\,\cclrd{s}$, uses the part specified by $\cclrd{r}$ to
  evaluate the function, ending with a value $\beta$ together with remaining context
  $\cclrd{s}$.
\item \textsc{CMTT interpretation.} The \ident{counit} operation evaluates a computation
  that requires no additional context to obtain a $\alpha$ value; given a function that
  turns a computation requiring context $\cclrd{r}$ into a value $\beta$, the \ident{cobind}
  operation can turn a computation that requires context $\cclrd{r}\,\cseq\,\cclrd{s}$
  into a computation that requires just $\cclrd{s}$ and contains $\beta$ (a part of the
  context requirements is eliminated by the function).
\end{itemize}

@@

The other aspect of practical implementation of coeffects is that of tracking the context requirements
(coeffect annotations) in the type system. To achieve this (without resorting to a single-purpose
language feature) the type system needs to be able to capture various kinds of coeffect algebras.
The structures required in this thesis include sets (with union or intersection), natural numbers (with
addition, maximum, minimum and multiplication), two-point lattice (for liveness) and free monoids
(vectors of annotations).

>>>>

\item In the flat coeffect calculus, we distinguish between \emph{splitting} context demands
  and \emph{merging} of context demands ($\cpar$ and $\czip$, respectively). In the structural
  coeffect calculus, the operations (which model appending and splitting vectors) are invertible
  and so the structural coeffect algebra uses the same tensor product $\atimes$. However, the
  $\cpar$ operation is still needed in the contraction rule.
\end{itemize}

\noindent
In the unified calculus presented in this section, we address the three differences as follows.
We include explicit rules for context manipulation in the calculus; in systems that arise from flat
calculi, the structural rules do not change coeffects and can thus be applied freely. We generalize
the structure of coeffect annotations using the notion of a ``container'' which can be specialized
to obtain a single annotation or a vector of annotations. This could potentially be used to capture
other structures such as trees in bunched typing \cite{substruct-bunched}. Finally, we distinguish
between splitting and merging of context demands (using the notation $\apar$ and $\azip$,
respectively). For structural coeffect calculi, the two operators coincide, but for flat
coeffect calculi, they differ and provide the needed flexibility.

@@

The $\cpar$ operation models combining of context demands arising from multiple parts of a
program. The meaning depends on the coeffect container. The operation can either combine requirements
of individual variables (structural coeffects) or requirements attached to the whole context of
multiple sub-exp\-ressions (flat coeffects).

@@

The difference between $\azip$ and $\apar$ is clarified by the semantics (Sec\-tion~\ref{sec:further-semantics}),
where $\aclrd{\textbf{r}}\,\azip\,\aclrd{\textbf{s}}$ is an annotation of the \emph{codomain} of
a morphism that merges the capabilities provided by two contexts (in the syntactic
reading, splits the context demands), while $\aclrd{\textbf{r}}\,\apar\,\aclrd{\textbf{s}}$
is an annotation of the \emph{domain} of a morphism that splits the capabilities of a single
context into two parts (in the syntactic reading, merges their context demands).
Syntactically, this means that we always use $\azip$ in the rule \emph{assumptions} and
$\apar$ in \emph{conclusions}.

@@

In the (\emph{app}), (\emph{abs}) and (\emph{let}) rules, the only change from the structural
system is that the vector concatenation $\atimes$ is now replaced with context splitting/merging
of the unified coeffect algebra. As already mentioned, we use splitting of context demands $\azip$
in rule \emph{assumptions} and merging of context demands $\apar$ in rule \emph{conclusions}.

@@

\noindent
Intuitively, the requirement $(\cclrd{r}\,\czip\,\cclrd{s}) \;\cleq\; (\cclrd{r}\,\cpar\,\cclrd{s})$,
which could be also written as $(\aclrd{r}\,\azip\,\aclrd{s}) \;\aclrd{\leq}\; (\aclrd{r}\,\apar\,\aclrd{s})$,
denotes that splitting context demands and then re-combining them preserves all the requirements
from the assumptions. The system may be imprecise and conclusions can overapproximate assumptions, but
it cannot lose requirements. This is fundamental for showing that exchange and contraction are admissible
in the unified system.

@@

\begin{itemize}
\item \textsc{Coeffect interpretation.} The \ident{counit} operation extracts a value
  and does not require any additional context; the \ident{cobind} operation requires
  context $\cclrd{r}\,\cseq\,\cclrd{s}$, uses the part specified by $\cclrd{r}$ to
  evaluate the function, ending with a value $\beta$ together with remaining context
  $\cclrd{s}$.
\item \textsc{CMTT interpretation.} The \ident{counit} operation evaluates a computation
  that requires no additional context to obtain a $\alpha$ value; given a function that
  turns a computation requiring context $\cclrd{r}$ into a value $\beta$, the \ident{cobind}
  operation can turn a computation that requires context $\cclrd{r}\,\cseq\,\cclrd{s}$
  into a computation that requires just $\cclrd{s}$ and contains $\beta$ (a part of the
  context demands is eliminated by the function).
\end{itemize}

@@

The other aspect of practical implementation of coeffects is that of tracking the context demands
(coeffect annotations) in the type system. To achieve this (without resorting to a single-purpose
language feature) the type system needs to be able to capture various kinds of coeffect algebras.
The structures required in this thesis include sets (with union or intersection), natural numbers (with
addition, maximum, minimum and multiplication), two-point lattice (for liveness) and free monoids
(vectors of annotations).

### 2016-05-29 text/backup/Chapter01.tex
Addressing feedback, fix demands (Chapter 6)

This example demonstrates how lambda abstraction behaves for context-dependent computations. The 
context requirement of the function body is a set of resources $\{\ident{Clock}, \ident{News}\}$.
The context requirements are split between the \emph{declaration site} and \emph{call site}.
However, there are multiple possible splittings. The splitting $\{\ident{News}\} \cup \{\ident{Clock}\}$
models the case when the database is accessed from the server, but time is taken from the client,
while the splitting $\{\ident{Clock}, \ident{News}\} \cup \{\}$ corresponds to the case when both
resources are accessed from the server.

@@

As an example, consider a language that allows us to get a value of a variable (representing
some changing data-source) \ident{x} versions back using the syntax $\pastval{a}{x}$. 
To track information about individual variables, we use a product-like operation $\times$ on tags 
to mirrors the product structure of variables. For example:
%
\begin{equation*}
\begin{array}{l}
(\ident{a}:\dnat, \ident{b}:\dnat) ~@~ 5 \times 10
  \vdash
    \pastval{a}{5}+\pastval{b}{10}: \ident{nat}
\end{array}
\end{equation*}
%
The annotations $5 \times 10$ corresponds to the free-variable context $\ident{a}, \ident{b}$, denoting
that we need at most 5 and 10 past values of \ident{a} and \ident{b}, respectively. If we substitute 
\ident{c} for both \ident{a} and \ident{b}, we need to combine the context-requirements and take the
maximum of the requirements of the individual variables:
%
\begin{equation*}
\begin{array}{l}
(\ident{c}:\dnat) ~@~ \ident{max}(5, 10)
  \vdash
    \pastval{c}{5}+\pastval{c}{10}: \ident{nat}
\end{array}
\end{equation*}
%
This version of the calculus removes the non-determinism of the lambda abstraction from the previous
version. As we associate information with individual variables, lambda abstraction creates a function
that requires the context requirements associated with the variable that is being abstracted over.

For example, if we wrapped the earlier example above in a function taking \ident{a} (and using 
\ident{b} from the declaration site) then the function would have context requirements $5$ -- that
is the number associated with the variable \ident{a}.

>>>>

This example demonstrates how lambda abstraction behaves for context-dependent computations. The 
context requirement of the function body is a set of resources $\{\ident{Clock}, \ident{News}\}$.
The context demands are split between the \emph{declaration site} and \emph{call site}.
However, there are multiple possible splittings. The splitting $\{\ident{News}\} \cup \{\ident{Clock}\}$
models the case when the database is accessed from the server, but time is taken from the client,
while the splitting $\{\ident{Clock}, \ident{News}\} \cup \{\}$ corresponds to the case when both
resources are accessed from the server.

@@

As an example, consider a language that allows us to get a value of a variable (representing
some changing data-source) \ident{x} versions back using the syntax $\pastval{a}{x}$. 
To track information about individual variables, we use a product-like operation $\times$ on tags 
to mirrors the product structure of variables. For example:
%
\begin{equation*}
\begin{array}{l}
(\ident{a}:\dnat, \ident{b}:\dnat) ~@~ 5 \times 10
  \vdash
    \pastval{a}{5}+\pastval{b}{10}: \ident{nat}
\end{array}
\end{equation*}
%
The annotations $5 \times 10$ corresponds to the free-variable context $\ident{a}, \ident{b}$, denoting
that we need at most 5 and 10 past values of \ident{a} and \ident{b}, respectively. If we substitute 
\ident{c} for both \ident{a} and \ident{b}, we need to combine the context-requirements and take the
maximum of the requirements of the individual variables:
%
\begin{equation*}
\begin{array}{l}
(\ident{c}:\dnat) ~@~ \ident{max}(5, 10)
  \vdash
    \pastval{c}{5}+\pastval{c}{10}: \ident{nat}
\end{array}
\end{equation*}
%
This version of the calculus removes the non-determinism of the lambda abstraction from the previous
version. As we associate information with individual variables, lambda abstraction creates a function
that requires the context demands associated with the variable that is being abstracted over.

For example, if we wrapped the earlier example above in a function taking \ident{a} (and using 
\ident{b} from the declaration site) then the function would have context demands $5$ -- that
is the number associated with the variable \ident{a}.
